Numerical simulations of the Cordilleran ice sheet through the last glacial cycle

### 2014-07-17 cordillera-cycle.tex
Polish section 4.2.3.

% =============================================================================
\section[Comparison to the geologic record]
        {Comparison to the geologic record\renote{
    Arjen has proposed: 2.1 Evaluation of model runs agains field evidence, 2.2
    Palaeoglaciology of the Cordilleran ice sheet, and 2.3 The last
    deglaciation; but I don't think that these headers work. In my view, this
    entire ``discussion'' part is all about evaluation of model runs agains
    field evidence and palaeoglaciology. It is necessary that this comparison
    starts here, in order to ``kick out'' the records that perform badly (ODP*,
    NGRIP) and justify that the rest of the analysis is performed on two
    (high-resolution) simulations only. Such figures as
    Figs.~\ref{fig:duration}--\ref{fig:erosion} and Fig.~\ref{fig:deglac},
    which I consider as different views of the results rather than new results
    (apart the fact that they use higher resolution) are drawn for the purpose
    of comparing model results against geological evidence. I propose to rename
    this section to a more explicit ``Comparison to the geologic record'' (or
    similar). I have also renamed the ``results'' to ``Sensitivity to
    atmospheric forcing time-series'', which is actually what that section is
    about: raw model output, no discussion against the geologic record. This
    new discussion is divided in three parts: 4.1 Glacial maxima, where we
    study the maximum stages, relatively well-documented by the geology; 4.2
    Nucleation centres, where we study ``intermediate'' states, for which less
    is known, but the model can say something; and 4.3 The last deglaciation,
    which deserves a separate section because many people are interested in
    this period, and much geological evidence is available.}}
\label{sec:discussion}
% =============================================================================

@@

It is generally believed that the Cordilleran ice sheet formed by the
coalescence of of several mountain-centred ice caps \citep{Davis.Mathews.1944}.
In our simulations, major ice-dispersal centres, visible on the modelled ice
cover duration maps (Fig.~\ref{fig:duration}), are located over the Coast
Mountains (CM), the Columbia and Rocky mountains (CR), the Skeena Mountains
(SM), and the Selwyn and MacKenzie Mountains (SMK). Although the Coast,
Columbia, Rocky and Skeena Mountains (CM, CR, SM) are covered by mountain
glaciers for most of the model time, providing durable nucleation centre for
the initiation of ice sheet growth, this is not the case for the Selwyn and
MacKenzie mountains (SMK), where ice cover on the highest peaks is limited to a
fraction of the last glacial cycle. In other words, the Selwyn and MacKenzie
mountains appear as a secondary ice-dispersal centre which ``activates'' only
during the coldest periods of the last glacial cycle.

@@

\todo{Explain why it would not be good with general statements on thermal
      evolution of ice cover.}

[In addition, the erosional landscape of the Skeena mountain bears one of the
strongest glacial imprint found within the area formerly covered by the
Cordilleran ice sheet \aref%
\footnote{\textbf{Arjen, Martin, Johan,} can we use a reference here to support
    this point (observations, mapping, shaded relief)? Alternatively, we could
    maybe use a photograph, if any of you has one. It would add a bit of
    variation into the figure list. \textbf{Johan}, you seemed to support this
    this point during our discussion, any idea?}.
Here we suggest that persistent ice cover (Fig.~\ref{fig:duration})
associated with basal ice temperatures at the pressure-melting point
(Fig.~\ref{fig:warmfrac} and ~\ref{fig:warmbase}%
\footnote{Fig.~\ref{fig:erosion} is tricky because it may be significantly
    affected by the constant-geothermal heat flux assumption, which is not
    valid for this region. Arjen, Johan and I decided to leave it for my kappa
    for now. Regarding Figs.~\ref{fig:warmfrac} and ~\ref{fig:warmbase}, I
    still wonder if it is useful to show both. Basically we have
    Fig.~\ref{fig:warmbase}\,=\,Fig.~\ref{fig:duration}\,*\,Fig.~\ref{fig:warmfrac}.
    Maybe the reader can perform this multiplication visually without need of
    Fig.~\ref{fig:warmbase}? Tell me what you think of it.})
may partly explain this spectacular landscape.]

A correlation is observed between the modelled duration of warm based ice cover
(Fig.~\ref{fig:warmbase}) and the degree of glacial modification of the
landscape (mainly in terms of the development of deep glacial valleys and
troughs). We find evidence for this on both the southwestern and northeastern
slopes of the southern Coast Mountains, on the western slopes of the Columbia
Mountains, on the western and eastern slopes of the northern Coast Mountains
and the Saint-Elias Mountains, and radiating off the Skeena Mountains
(Figs.~\ref{fig:duration}, \ref{fig:warmbase} and~X\renote{
    Yes it could be interesting to compare basal flow patterns with Fig.~2 of
    \citet{Kleman.etal.2010}, but I don't think that I have the time for this
    right now.}).
The Skeena Mountains, for example, indeed bear a strong glacial imprint that
indicates ice drainage in a system of distinct glacial troughs emanating in a
radial pattern from the centre of the mountain range (Fig. X). We suggest that
persistent ice cover (Fig.~\ref{fig:duration}) associated with basal ice
temperatures at the pressure-melting point (Figs.~\ref{fig:warmfrac}
and~\ref{fig:warmbase}) explains the large-scale glacial erosional imprint on
the landscape. In contrast, a well-developed network of glacial valleys west of
the MacKenzie Mountains \citep[Fig.~X;][]{Kleman.etal.2010,
Stroeven.etal.2010}, which is modelled to have hosted warm-based ice
(Fig.~\ref{fig:warmfrac}) has only been glaciated for a short duration
(Fig.~\ref{fig:duration}) during the last glacial cycle according to our
results (Figs.~\ref{fig:duration} and~\ref{fig:warmbase}). We therefore infer
that either our model does not perform reliably in this part of the domain or
the observed landscape pattern originates from older glacial cycles and
indicates an increased relative importance of this ice dispersal centre prior
to the Late Pleistocene \citep[cf.][]{Ward.etal.2008, Demuro.etal.2012}.

\todo{Paragraph on model uncertainties related to warm-based ice.}

@@

% fig:duration
\begin{figure*}
  \includegraphics{duration}
  \caption{Modelled duration of ice cover during the last 120\,kyr.
           Note the irregular colour scale. A contiguous ice cover spanning
           from the Alaska Range (AR) to the southern Coast Mountains (CM) and
           Rocky Mountains (RM) exists for about 29\,kyr in both
           simulations. A central
           ice cover persists over the Skeena Mountains (SM) during most of the
           simulation. On the other hand, the maximal extent of the ice sheet
           generally corresponds to relatively short durations of ice cover.}
  \label{fig:duration}
\end{figure*}

@@

% fig:erosion
\begin{figure*}
  \includegraphics{erosion}
  \caption{Modelled cumulative basal displacement (integrand of basal velocity)
           over the last 120\,kyr.
           \todo{remove this figure (see footnote in text).}}
  \label{fig:erosion}
\end{figure*}

>>>>

% =============================================================================
\section[Comparison to the geologic record]
        {Comparison to the geologic record\renote{
    Arjen has proposed: 2.1 Evaluation of model runs agains field evidence, 2.2
    Palaeoglaciology of the Cordilleran ice sheet, and 2.3 The last
    deglaciation; but I don't think that these headers work. In my view, this
    entire ``discussion'' part is all about evaluation of model runs agains
    field evidence and palaeoglaciology. It is necessary that this comparison
    starts here, in order to ``kick out'' the records that perform badly (ODP*,
    NGRIP) and justify that the rest of the analysis is performed on two
    (high-resolution) simulations only. Such figures as
    Figs.~\ref{fig:duration}--\ref{fig:warmfrac} and Fig.~\ref{fig:deglac},
    which I consider as different views of the results rather than new results
    (apart the fact that they use higher resolution) are drawn for the purpose
    of comparing model results against geological evidence. I propose to rename
    this section to a more explicit ``Comparison to the geologic record'' (or
    similar). I have also renamed the ``results'' to ``Sensitivity to
    atmospheric forcing time-series'', which is actually what that section is
    about: raw model output, no discussion against the geologic record. This
    new discussion is divided in three parts: 4.1 Glacial maxima, where we
    study the maximum stages, relatively well-documented by the geology; 4.2
    Nucleation centres, where we study ``intermediate'' states, for which less
    is known, but the model can say something; and 4.3 The last deglaciation,
    which deserves a separate section because many people are interested in
    this period, and much geological evidence is available.}}
\label{sec:discussion}
% =============================================================================

@@

It is generally believed that the Cordilleran ice sheet formed by the
coalescence of of several mountain-centred ice caps \citep{Davis.Mathews.1944}.
In our simulations, major ice-dispersal centres, visible on the modelled ice
cover duration maps (Fig.~\ref{fig:duration}), are located over the Coast
Mountains (CM), the Columbia and Rocky mountains (CR), the Skeena Mountains
(SM), and the Selwyn and MacKenzie Mountains (SMK). Although the Coast,
Columbia, Rocky and Skeena Mountains (CM, CR, SM) are covered by mountain
glaciers for most of the model time, providing durable nucleation centre for
the initiation of ice sheet growth, this is not the case for the Selwyn and
MacKenzie mountains (SMK), where ice cover on the highest peaks is limited to a
fraction of the last glacial cycle. In other words, the Selwyn and MacKenzie
mountains appear as a secondary ice-dispersal centre which ``activates'' only
during the coldest periods of the last glacial cycle. The Northern Rocky
Mountains (Fig.~\ref{fig:duration}, NR), do not act as a nucleation centre,
but rather as a pinning point for the ice margin coming from the west.

@@

A correlation is observed between the modelled duration of warm based ice cover
(Fig.~\ref{fig:warmbase}) and the degree of glacial modification of the
landscape (mainly in terms of the development of deep glacial valleys and
troughs). We find evidence for this on both the southwestern and northeastern
slopes of the southern Coast Mountains, on the western slopes of the Columbia
Mountains, on the western and eastern slopes of the northern Coast Mountains
and the Saint-Elias Mountains, and radiating off the Skeena Mountains
(Figs.~\ref{fig:duration} and \ref{fig:warmbase};
\citealp[Fig.~2]{Kleman.etal.2010})\renote{
    Yes, I agree that it would be good to eventually reproduce here the basal
    flow patterns from Johan's map \citep[Fig.~2]{Kleman.etal.2010}, but I
    don't think that I will have the time for this before thesis submission}.
The Skeena Mountains, for example, indeed bear a strong glacial imprint that
indicates ice drainage in a system of distinct glacial troughs emanating in a
radial pattern from the centre of the mountain range
\citep[Fig.~2]{Kleman.etal.2010}. We suggest that
persistent ice cover (Fig.~\ref{fig:duration}) associated with basal ice
temperatures at the pressure-melting point (Figs.~\ref{fig:warmfrac}
and~\ref{fig:warmbase}) explains the large-scale glacial erosional imprint on
the landscape. A well-developed network of glacial valleys west of
the MacKenzie Mountains (\citealp[Fig.~2]{Kleman.etal.2010}; \citealp[Fig.~8]
{Stroeven.etal.2010}) is modelled to have hosted warm-based ice
(Fig.~\ref{fig:warmfrac}) but it has only been glaciated for a short fraction
of the last glacial cycle (Fig.~\ref{fig:duration}) according to our
results. This perhaps indicate that\renote{
    I don't think that we have to infer that ``our model does not perform
    reliably'' here. I am actually quite happy to discover that the model
    puts warm-based ice where there is (another) region of strong glacial
    erosion. I would rather say that computing the ``duration of
    warm-based ice cover'' as I did for is a very simple approach and that we
    can expect it to give a qualitative proxy for glacial erosion but not a
    quantitative one.}
the observed landscape pattern originates from older glacial cycles and
witnesses an increased relative importance of this ice dispersal centre prior
to the Late Pleistocene \citep[cf.][]{Ward.etal.2008, Demuro.etal.2012}.

The modelled distribution of warm-based ice cover (Figs.~\ref{fig:warmbase}
and~\ref{fig:warmfrac}) is inevitably affected by our assumption of a constant,
70\,\unit{mW\,m^{-2}} geothermal heat flux at 3\,km depth
(Sect.~\ref{sec:overview}). However, both the Skeena Mountain area and the area
west of the MacKenzie Mountains are areas of higher-than-average geothermal
heat flux \citep{Blackwell.Richards.2004}. We can therefore expect a higher
frequency of warm-based ice cover in this areas if we were to include
spatially-variable geothermal forcing in our simulations.

@@

% fig:duration
\begin{figure*}
  \includegraphics{duration}
  \caption{Modelled duration of ice cover during the last 120\,ka.
           Note the irregular colour scale. A contiguous ice cover spanning
           from the Alaska Range (AR) to the southern Coast Mountains (CM) and
           Rocky Mountains (RM) exists for about 29\,ka in both
           simulations. A central
           ice cover persists over the Skeena Mountains (SM) during most of the
           simulation. On the other hand, the maximal extent of the ice sheet
           generally corresponds to relatively short durations of ice cover.
           \todo{Add WSE, SMK, NR, NC mountain ranges, and changes kyr to ka}}
  \label{fig:duration}
\end{figure*}

@@
